---
title: "Single Cell Analysis Workshop"
author:
- Jonathan Ronen
- Vedran Franke
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
    number_sections: yes
  html_notebook:
    number_sections: yes
    self_contained: yes
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache   = TRUE,
                      include = FALSE, 
                      eval    = FALSE)

```


# Intro

Through this tutorial you should get familiar with the structure of the 
SingleCellExperiment object, and basic functions for manipulation, 
and quality control of the single cell expression data. 


## Single Cell Experiment Object

SingleCellExperiment object is a container for single cell data.
It is build upon the SummarizedExperiment object, with the following structure

![SingleCellExperiment](SingleCellExperiment.svg)




## The Data

For this workshop we will use the publicly available data set, provided by
10x genomics.
The count matrix can be found on the following link:
[link](https://s3-us-west-2.amazonaws.com/10x.files/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz).

We have pre-processed the data into a hdf5 file. **hdf5** file format is constructed
for fast manipulation with array based data types (i.e. matrices). It stores
indexed location for fast random access. You can consider the hdf5 file like
a 2d bam file.


## Preliminaries:

If you haven't installed the required R packages, please install them now:
```{r install, eval=FALSE, include=TRUE}
source("https://bioconductor.org/biocLite.R")

biocLite(c("data.table","DelayedArray","DelayedMatrixStats","dplyr","GenomicAlignments","GenomicFiles","GenomicRanges","HDF5Array","pheatmap","Rsamtools","rtracklayer","scater","scran","SingleCellExperiment","stringr","yaml", "devtools"))

devtools::install_github("csoneson/iSEE", ref='3.4')

devtools::install_github("bimsbbioinfo/netSmooth", ref='bioconductor_3_6')

```

In addition, please install the following packages:

ggplot2, biomaRt, cowplot, ComplexHeatmap, dplyr

```{r install2, includ=TRUE, eval=FALSE}
source("https://bioconductor.org/biocLite.R")

biocLite(c('ggplot2','biomaRt','cowplot','ComplexHeatmap','dplyr'))
```




# Exercises

### Constructing SingleCellExperiment Objects

From the files in the data folder construct the SingleCellExperiment object.
Hint: h5ls, HDF5Array, SingleCellExperiment

First you will need to connect to the hdf5 object. Then you will
pass the object to the SingleCellExperiment constructor.

To connect to the hdf5 file, you will need the name of the saved matrix.
This information can be obtained using the h5ls function

```{r sce_construct, cache=TRUE}
suppressPackageStartupMessages(library(HDF5Array))
input_hdf5 = 's10x_3cells.hdf5'
name = h5ls(input_hdf5)$name

mat = HDF5Array(file=input_hdf5, name=name)

suppressPackageStartupMessages(library(SingleCellExperiment))
sce = SingleCellExperiment(
     assays  = list(counts=mat))

```

### What is the size of the constructed single cell experient?

```{r sce_size}
print(object.size(sce),units='Mb')
```
### What do the following functions do?

assays

rowData

colData


### Add the gene names, and biotypes to the SingleCellExperiment object

The table **genes.tsv** contains ENSEMBL gene ids for the corresponding 
rows of the expression matrix. Load the table into R.

Using the biomaRt package, download the gene names and gene biotypes
for the genes in the SingleCellExperiment object, and update the gene annotations.
Hint: useMart, getBM, rowData

```{r gene_names}
gene_names = read.table('genes.tsv')

library(biomaRt)
mart = useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl")
bm   = getBM(mart       = mart,
             attributes = c('ensembl_gene_id','hgnc_symbol','gene_biotype'),
             filters    = 'ensembl_gene_id',
             values     = gene_names$V1)

bm = bm[match(gene_names$V1, bm$ensembl_gene_id),]

rowData(sce) = DataFrame(bm)


head(rowData(sce))
```

### Remove all genes which could not be annotated in the previous step

Remove all rows with NA values.

```{r remove_rows}
sce = sce[!is.na(rowData(sce)$ensembl_gene_id)]
```


# SingleCellExperiment Exploration

In this step we will learn how to extract the basic information from 
the SingleCellExperiment object.
For this we will use function from the DelayedMatrixStats package.
DelayedMatrixStats package

### How many lincRNAs are analyzed in the data?

```{r, lincRNA}
sum(rowData(sce)$gene_biotype == 'lincRNA')
```


### Remove all genes which have less than 100 reads in all cells

Hint: DelayedMatrixStats package

```{r filter_genes}
suppressPackageStartupMessages(library(DelayedMatrixStats))
rs = DelayedMatrixStats::rowSums2(assays(sce)$counts)
sce = sce[rs > 100,]
```
